---
title: "Missing Data Tutorial"
author: "Yichi Zhang"
date: "2023-03-30"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# install.packages("mice")
# install.packages("naniar")
library(mice)
library(tidyverse)
library(naniar)
library(rio)
```


```{r}
# import the linelist
linelist <- import("linelist_cleaned.rds")
# view the first 6 observations of the dataset
head(linelist)
```

## Problems with Missingness

Sometimes you might encounter the following errors when you started analyzing your data...

```{r}
mean(linelist$age)
```

<!-- ```{r} -->
<!-- lm(bmi ~ age + gender + wt_kg + ht_cm, data = linelist, na.action = na.pass) -->
<!-- ``` -->

These functions could work after dealing with missing data. In R, NA represents all types of missing data.

```{r, results='hide'}
mean(linelist$age, na.rm = TRUE)
# use na.omit to exclude incomplete cases, sample size will change
mod_omit <- lm(bmi ~ age + gender + wt_kg + ht_cm, data = linelist, na.action = na.omit)
resid(mod_omit)
# use na.exclude to exclude incomplete cases in the analysis but still keep the them in the dataset
mod_exclude <- lm(bmi ~ age + gender + wt_kg + ht_cm, data = linelist, na.action = na.exclude)
resid(mod_exclude)
# could check the help page of na.action to see how R handles missing values
# ? na.action
```

## Assess Missingness

The number of NAs in the dataset by columns.

```{r}
# check the number of NAs in the dataset by columns
colSums(is.na(linelist))
# the dimension of the original dataset
dim(linelist)
```
The proportion of missing values in each variable. 

```{r}
linelist %>%
    # check each variable's missing values
    map(is.na) %>%
    # calculate the total sum of missing values in each variable
    map(sum) %>%
    # pick the sum of missing values in each variable and divide by the sample size
    map(~ . /nrow(linelist))%>%
    # bind multiple columns together
    bind_cols()
```
We can compute missingness in different ways.

```{r}
# percent of ALL data frame values that are missing
pct_miss(linelist)
# percent of rows with any value missing
pct_miss_case(linelist)
# percent of rows that are complete
pct_complete_case(linelist)
```

### Functions to Remove Missing Values

```{r, results='hide'}
## listwise deletion
na.omit(linelist)
## drops rows missing values for any of these columns
linelist %>%
  drop_na(case_id, date_onset, age)
## remove NAs by specifing the input parameters
mean(linelist$age, na.rm = TRUE)
## complete cases
linelist[complete.cases(linelist), ]
```

## Visualize Missingness

We can use `vis_miss()` to visualize the dataframe as a heatmap, showing whether each value is missing or not.

```{r}
# check missing data pattern for variables age, temperature, and fever
md.pattern(linelist[, c("age", "temp", "fever")])
## show the number of missing in each column
gg_miss_var(linelist, show_pct = TRUE)
## split the data by a variable
linelist %>% 
  gg_miss_var(show_pct = TRUE, facet = outcome)
```
In the scatterplot below, the red dots are records where the value for one column is present but the value for the other column is missing. This allows you to see the distribution of missing values in relation to the non-missing values.

```{r}
ggplot(
  data = linelist,
  mapping = aes(x = age_years, y = temp)) +     
  geom_miss_point()
```

`gg_miss_fct()` assesses missingness in the data frame stratified by another column, which returns a heatmap of percent missingness in the data frame by a factor/categorical (or date) column. 

```{r}
gg_miss_fct(linelist, age_cat5)
## change over time
gg_miss_fct(linelist, date_onset)
```

Here we take the linelist, add a new column for week, group the data by week, and then calculate the percent of that week’s records where the value is missing. Then we plot the proportion missing as a line, by week. 

```{r}
outcome_missing <- linelist %>%
    mutate(week = lubridate::floor_date(date_onset, "week")) %>% # create new week column
    group_by(week) %>%                                           # group the rows by week
    summarise(                                                   # summarize each week
        n_obs = n(),                                             # number of records
        outcome_missing = sum(is.na(outcome) | outcome == ""),   # number of records missing the value
        outcome_p_miss  = outcome_missing / n_obs,               # proportion of records missing the value
        outcome_dead    = sum(outcome == "Death", na.rm=T),      # number of records as dead
        outcome_p_dead  = outcome_dead / n_obs) %>%              # proportion of records as dead
    tidyr::pivot_longer(-week, names_to = "statistic") %>%       # pivot all columns except week, to long format for ggplot
    filter(stringr::str_detect(statistic, "_p_"))                # keep only the proportion values

ggplot(data = outcome_missing)+
    geom_line(
      mapping = aes(x = week, y = value, group = statistic, color = statistic),
      size = 2,
      stat = "identity")+
    labs(title = "Weekly outcomes",
         x = "Week",
         y = "Proportion of weekly records") + 
     scale_color_discrete(
       name = "",
       labels = c("Died", "Missing outcome"))+
    scale_y_continuous(breaks = c(seq(0,1,0.1)))+
  theme_minimal()+
  theme(legend.position = "bottom")
```


## Address Missingness

### Mean Imputation

```{r, results='hide'}
linelist[is.na(linelist$temp),]
linelist <- linelist %>%
    mutate(temp_replace_na_with_mean = replace_na(temp, mean(temp, na.rm = T)))
linelist[is.na(linelist$temp),]
```

### Regression Imputation

```{r}
simple_temperature_model_fit <- lm(temp ~ fever + age_years, data = linelist)
#using our simple temperature model to predict values just for the observations where temp is missing
predictions_for_missing_temps <- predict(simple_temperature_model_fit,
                                         newdata = linelist %>% filter(is.na(temp)))
```

```{r}
model_dataset <- linelist %>%
    select(temp, fever, age_years)  
temp_imputed <- mice(model_dataset,
                     # linear regression, predicted values
                     method = "norm.predict",
                     # set seed for reproducibility 
                     seed = 1,
                     # number of multiple imputations, default 5
                     m = 1,
                     print = F)
```

### Multiple Imputation

```{r}
# imputing missing values for all variables in our model_dataset, and creating 10 new imputed datasets
multiple_imputation = mice(
    model_dataset,
    seed = 1,
    m = 10,
    print = FALSE) 
# inspect the regression model with the imputed data
model_fit <- with(multiple_imputation, lm(temp ~ age_years + fever))
base::summary(mice::pool(model_fit))
pool(model_fit)
plot(multiple_imputation)
```
The fraction of information missing due to nonresponse (fmi) and the relative increase in variance due to nonresponse (lambda) are pretty high. 

```{r}
## imputed datasets
multiple_imputation
# information stores in the object multiple_imputation
attributes(multiple_imputation)
# original dataset
# multiple_imputation$data
# imputated dataset
# multiple_imputation$imp
# extract a certain imputed dataset
c3 <- complete(multiple_imputation, 1)
```


## Exercise

Explore the following dataset `airquality`

```{r, results='hide'}
airquality
```

